Measurement of ($\pi^-$-Ar) and ($K^+$-Ar) total hadronic cross sections at the LArIAT experiment

### 2018-06-25 Chapter-2/2-LArTPCs.tex
s

\section{The Liquid Argon Time Projection Chamber Technology}
In this section, we outline an extremely brief history of Time Projection Chambers as particle detectors, focusing on their incarnation as Argon detectors for neutrino physics. We further describe the working principles of Liquid Argon Time Projection Chambers, leading  to the description of the event reconstruction in LArTPC.

@@

Excitation and ionization of the detector medium occur in similar amounts. Since the ionizing collisions occur randomly, we can parametrize their number $k$ in a segment of length $s$ along the track  with a Poissonian function
\begin{equation}
P(k) = \frac{s^k}{k! \lambda^k} e^{-s/ \lambda }, 
\end{equation}
where $\lambda = 1/N_{e}\sigma_i$, with $N_{e}$ being the electron density of $\sigma_i$ the ionization cross-section per electron.  About 66\% of the ionizing collisions in Argon produce only a single electron/ion pair \cite{0034-4885-73-11-116201}; in the other cases, the transferred kinetic energy is enough for the primary electron to liberate one or more secondary electrons, which usually stay close to the original pair.  
Occasionally, electrons can receive enough energy to be ejected with high energy, forming a so-called ``$\delta$-ray": a detectable  short track off the particle trajectory, as shown in figure \ref{fig:delta}. 
The average number of $\delta$-ray  with energy E$>$E$_0$ per cm follows the empirical form
\begin{equation}
P(E>E_0) \sim \frac{y}{\beta^2 E_0},
\end{equation}
where $y$ is an empirical factor depending on the medium (0.114 for  gaseous Ar), and $\beta$ is $v/c$.

\begin{figure}[hbpt]
\centering
\includegraphics[width=\textwidth]{Chapter-2/Images/Delta.png}
\caption{Events display for a LArIAT pion absorption candidate on the induction plane, with highlighted delta ray.}
\label{fig:delta}
\end{figure}

@@

\textbf{Hit Reconstruction.} The second stage of the signal processing is the reconstruction of hits, indicating an energy deposition in the detector.  A peak finder scans the deconvolved TPC waveforms for each wire on the whole readout time looking for spikes  above  the waveform's baseline. It then fits these peaks with gaussian shapes and stores the fit parameters such as the quality of the fit, the peak time, height and area under the gaussian fit. The information resulting from this process on a single spike form a single reconstructed ``hit", see Figure \ref{fig:SignalProc} panel c).
The next steps in the event reconstruction chain will then decide if rejecting hits with poor fits.
It is important to notice how the height and width of the hit depend on the topology of the event: for example, a particle running  parallel to the wire planes will leave a series of narrow individual hits, one on each consecutive wire, while a particle traveling towards the planes will leave long, wide hits on very few wires. The height of the hits and their integral is proportional to the charge collected on the wire, so it depends on the particle type.\\

The event reconstruction chain uses collection of hits to form more complex objects associated with the particles in the detector. The development of different approaches to accomplish this task is an extremely hot topic in LArTPC event reconstruction which spans from more traditional approaches such as line-clustering  \cite{Barker2011} to the use of machine learning tools \cite{1748-0221-12-03-P03011}. Generally speaking, the scope of hit clustering and event reconstruction to provide shower-like or track like-objects with an associated energy reconstruction. This is because different particles have different topology in the detector -- electrons and photon create electromagnetic showers,  resulting in shower-like topologies, while muons and hadrons  leave track-like signals.  For the scope of these thesis, we will describe only LArIAT's approach to track reconstruction even if we recognize the breath of LArTPC event reconstruction is much wider. We are interested in the reconstruction of pions and kaons in the active volume, whose topology is track-like.\\

@@

\textbf{Calorimetry.} The last step in the event reconstruction chain is to assign calorimetric information to the track (or shower) objects. Calorimetry is performed separately on the different planes. A multi-step procedure is needed to retrieve the energy deposited in the TPC  from the charge seen by the wires.
For each hit associated with the track object, the calorimetry algorithms calculate the charge seen on every wire using the area underneath the gaussian fit; then, they correct this raw charge by the electron life time, the electronic noise on the considered wire and the recombination effect. Lastly an overall calibration of the energy, explained in detail in section \ref{ch:energyCalibration}, is applied and the calorimetric information for the given track is assigned.
Even if calorimetry is done in 2D, it benefits from the 3D tracking information; typical objects available after the calorimetric reconstruction are the total energy deposited by the particle and its stopping power $dE/dx$ at each ``track pitch", i.e. at each 2D projection on the wire plane of the 3D trajectory.

>>>>

\section{The Liquid Argon Time Projection Chamber Technology}
In this section, we outline an extremely brief history of Time Projection Chambers as particle detectors, focusing on their incarnation as argon detectors for neutrino physics. We further describe the working principles of Liquid Argon Time Projection Chambers, leading  to the description of the event reconstruction in LArTPC.

@@

Excitation and ionization of the detector medium occur in similar amounts. Since the ionizing collisions occur randomly, we can parametrize their number $k$ in a segment of length $s$ along the track  with a Poissonian function
\begin{equation}
P(k) = \frac{s^k}{k! \lambda^k} e^{-s/ \lambda }, 
\end{equation}
where $\lambda = 1/N_{e}\sigma_i$, with $N_{e}$ being the electron density of $\sigma_i$ the ionization cross-section per electron.  About 66\% of the ionizing collisions in argon produce only a single electron/ion pair \cite{0034-4885-73-11-116201}; in the other cases, the transferred kinetic energy is enough for the primary electron to liberate one or more secondary electrons, which usually stay close to the original pair.  
Occasionally, electrons can receive enough energy to be ejected with high energy, forming a so-called ``$\delta$-ray": a detectable  short track off the particle trajectory, as shown in figure \ref{fig:delta}. 
The average number of $\delta$-ray  with energy E$>$E$_0$ per cm follows the empirical form
\begin{equation}
P(E>E_0) \sim \frac{y}{\beta^2 E_0},
\end{equation}
where $y$ is an empirical factor depending on the medium (0.114 for  gaseous Ar), and $\beta$ is $v/c$.

\begin{figure}[hbpt]
\centering
\includegraphics[width=\textwidth]{Chapter-2/Images/Delta.png}
\caption{Event display for a LArIAT pion absorption candidate on the induction plane, with highlighted delta ray.}
\label{fig:delta}
\end{figure}

@@

\textbf{Hit Reconstruction.} The second stage of the signal processing is the reconstruction of hits, indicating an energy deposition in the detector.  A peak finder scans the deconvolved TPC waveforms for each wire on the whole readout time looking for spikes  above  the waveform's baseline. It then fits these peaks with gaussian shapes and stores the fit parameters such as the quality of the fit, the peak time, height and area under the gaussian fit. The information resulting from this process on a single spike form a single reconstructed ``hit", see Figure \ref{fig:SignalProc} panel c).
The next steps in the event reconstruction chain will then decide which hits to use according to their goodness of fit. It is important to notice how the height and width of the hit depend on the topology of the event: for example, a particle running  parallel to the wire planes will leave a series of narrow individual hits, one on each consecutive wire, while a particle traveling towards the planes will leave long, wide hits on very few wires. The height of the hits and their integral is proportional to the charge collected on the wire, so it depends on the particle type.\\

The event reconstruction chain uses collection of hits to form more complex objects associated with the particles in the detector. The development of different approaches to accomplish this task is an extremely hot topic in LArTPC event reconstruction which spans from more traditional approaches such as line-clustering  \cite{Barker2011} to the use of machine learning tools \cite{1748-0221-12-03-P03011}. Generally speaking, the scope of hit clustering and event reconstruction is to provide shower-like or track like-objects with an associated energy reconstruction. This is because different particles have different topology in the detector -- electrons and photon create electromagnetic showers,  resulting in shower-like topologies, while muons and hadrons  leave track-like signals.  For the scope of this thesis, we will describe only LArIAT's approach to track reconstruction even if we recognize the breath of LArTPC event reconstruction is much wider. We are interested in the reconstruction of pions and kaons in the active volume, whose topology is track-like.\\

@@

\textbf{Calorimetry.} The last step in the event reconstruction chain is to assign calorimetric information to the track (or shower) objects. Calorimetry is performed separately on the different planes. A multi-step procedure is needed to retrieve the energy deposited in the TPC  from the charge seen by the wires.
For each hit associated with the track object, the calorimetry algorithms calculate the charge seen on every wire using the area underneath the gaussian fit; then, they correct this raw charge by the electron life time, the electronic noise on the considered wire and the recombination effect. Lastly an overall calibration of the energy, explained in detail in section \ref{ch:energyCalibration}, is applied and the calorimetric information for the given track is assigned.
Even if calorimetry is done in 2D, it benefits from the 3D tracking information; typical objects available after the calorimetric reconstruction are the total energy deposited by the particle and its stopping power $dE/dx$ at each ``track pitch", i.e. at each segment between two 3D points projected on the wire plane.

### 2018-06-25 Chapter-3/3-TheLArIATExperiment.tex
s

\begin{figure}[!h]
\begin{centering}
\vspace{-0.3cm}
\includegraphics[height=3.0in]{Chapter-3/Images/ExcitationCurves.png}
\caption{
{ Magnetic field over current as a function of the current, for one NDB magnet (excitation curve). The data was collected using two hall probes (blue and green). We fit the readings with a cubic function (black) to average of measurements (red) given in the legend \cite{detectorPaper}.}
}
\label{fig:magnet_excitation}
\end{centering}
\end{figure}

The current through the magnets at a given time is identical in both magnets. For the Run II data taking period, the current settings explored were 60A (B $\sim$0.21 T) and 100A (B $\sim$0.35 T) in both polarities. 
Albeit advantageous to enrich the tertiary beam composition with high mass particles such as kaons, we never pushed the magnets current over 100 A, so as not to incur in overheating.  During operation, we operated an air and water cooling system on the magnets and we remotely monitored the magnet temperatures.

\subsection{Multi-Wire Proportional Chambers}\label{sec:MWPC}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\begin{centering}
\vspace{-0.3cm}
\includegraphics[height=2.3in]{Chapter-3/Images/WireChamber.png}
\caption{
{One of the four Multi Wire Proportional Chambers (WC) used in the LArIAT tertiary beamline and related read-out electronics.}
}
\label{fig:wirechamber}
\end{centering}
\end{figure}

@@

Two ASDQ chips~\cite{ASDQchip} mounted on a mother board plugged into the chamber serve as front end amplifier/discriminator. The chips are connected to a multi-hit TDC~\cite{Sten} which provides a fast OR output used as first level trigger. The TDC time resolution is 1.18~ns/bin and can accept 2 edges per 9~ns.  
The maximum event rate acceptable by the chamber system is 1 MHz.%: this rate is not a limiting factor considering that the rate of the tertiary particle beam at the first wire chamber is estimated to be less than 15 kHz}. 
A full spill of data occurring once per supercycle is stored on the TDC board memory at once and read out by a specially designed controller.  We use LVDS cables to carry both power and data between the controller and the TDCs and from the controller to the rest of the DAQ.  
%It is possible to program the time window for acceptance for hits, time offsets, front end threshold, and pulse shaping parameters through the controller via a USB from a PC or through an Ethernet connection.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{LArIAT Cosmic Ray Paddle Detectors}\label{sec:CosmicRayPaddle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
LArIAT triggers both on beam events and on cosmic rays events. We perform this latter trigger by using two sets of cosmic ray paddle detectors (a.k.a. ``cosmic towers".) The cosmic towers frame the LArIAT cryostat, as one sits in the downstream left corner and the other sits in the upstream right corner of the cryostat. Two paddle sets of four scintillators pieces each make up each cosmic tower, an upper set and a lower set per tower. 
Of the four paddles, a couple of two matched paddles stands upright while the a second matched pair lies across the top of the assembly in the top sets (or across the bottom of the assembly in the bottom sets). The horizontal couple is used as a veto for particles traveling from inside the TPC out.  The four signals  from the vertical paddles along one of the body diagonals of the TPC are combined in a logical ``AND''. This allows to select track due to cosmic muons at the ground level crossing the TPC along one of its diagonals.  Cosmic ray muons whose average energy is in the few GeV range crossing both anode and cathode populate the events triggered this way. This particularly useful sample of tracks is associated can be used for many tasks; for example, we use anode-cathode piercing tracks to cross check the TPC electric field on data (see Appendix \ref{ch:AppendixB}), to calibrate the charge response of the TPC wires for the full TPC volume and to measure the electron lifetime in the chamber \cite{LArIATLifeTime}.

>>>>

\begin{figure}[!h]
\begin{centering}
\vspace{-0.3cm}
\includegraphics[height=3.0in]{Chapter-3/Images/ExcitationCurves.png}
\caption{
{ Magnetic field over current as a function of the current, for one NDB magnet (excitation curve). The data were collected using two hall probes (blue and green). We fit the readings with a cubic function (black) to average of measurements (red) given in the legend \cite{detectorPaper}.}
}
\label{fig:magnet_excitation}
\end{centering}
\end{figure}

The current through the magnets at a given time is identical in both magnets. For the Run II data taking period, the current settings explored were 60A (B $\sim$0.21 T) and 100A (B $\sim$0.35 T) in both polarities. 
Albeit advantageous to enrich the tertiary beam composition with high mass particles such as kaons, we never pushed the magnets current over 100 A, so to not overheat them.  During operation, we operated an air and water cooling system on the magnets and we remotely monitored the magnet temperatures.

\subsection{Multi-Wire Proportional Chambers}\label{sec:MWPC}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\begin{centering}
\vspace{-0.3cm}
\includegraphics[height=2.3in]{Chapter-3/Images/WireChamber.png}
\caption{
{One of the four Multi-Wire Proportional Chambers (WC) used in the LArIAT tertiary beamline and related read-out electronics.}
}
\label{fig:wirechamber}
\end{centering}
\end{figure}

@@

Two ASDQ chips~\cite{ASDQchip} mounted on a mother board plugged into the chamber serve as front end amplifier/discriminator. The chips are connected to a multi-hit TDC~\cite{Sten} which provides a fast OR output used as first level trigger. The TDC time resolution is 1.18~ns/bin and can accept 2 edges per 9~ns.  
The maximum event rate acceptable by the chamber system is 1 MHz. 
A full spill of data occurring once per supercycle is stored on the TDC board memory at once and read out by a specially designed controller.  We use LVDS cables to carry both power and data between the controller and the TDCs and from the controller to the rest of the DAQ.  
%It is possible to program the time window for acceptance for hits, time offsets, front end threshold, and pulse shaping parameters through the controller via a USB from a PC or through an Ethernet connection.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{LArIAT Cosmic Ray Paddle Detectors}\label{sec:CosmicRayPaddle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
LArIAT triggers both on beam events and on cosmic rays events. We perform this latter trigger by using two sets of cosmic ray paddle detectors (a.k.a. ``cosmic towers".) The cosmic towers frame the LArIAT cryostat, as one sits in the downstream left corner and the other sits in the upstream right corner of the cryostat. Two paddle sets of four scintillators pieces each make up each cosmic tower, an upper set and a lower set per tower. 
Of the four paddles, a couple of two matched paddles stands upright while the a second matched pair lies across the top of the assembly in the top sets (or across the bottom of the assembly in the bottom sets). The horizontal couple is used as a veto for particles traveling from inside the TPC out.  The four signals  from the vertical paddles along one of the body diagonals of the TPC are combined in a logical ``AND''. This allows to select track due to cosmic muons at the ground level crossing the TPC along one of its diagonals.  Cosmic ray muons whose average energy is in the few GeV range crossing both anode and cathode populate the events triggered this way. This particularly useful sample of tracks can be used for many tasks; for example, we use anode-cathode piercing tracks to cross check the TPC electric field on data (see Appendix \ref{ch:AppendixB}), to calibrate the charge response of the TPC wires for the full TPC volume and to measure the electron lifetime in the chamber \cite{LArIATLifeTime}.

### 2018-06-25 Chapter-4/4-Methodology.tex
s

This chapter describes the general procedure employed to measure  total hadronic interaction cross sections on Ar in LArIAT.
Albeit with small differences, both the  ($\pi^{-}$,Ar) and (K$^{+}$,Ar) total hadronic cross section measurements rely on the same procedure. We start by selecting the particle of interest using a combination of beamline detectors and TPC information (Section \ref{ch:ParticleSelectionMethod}). We then perform a handshake between the beamline information and the TPC tracking to assure the selection of the correct TPC track (Section \ref{ch:WC2TPCMatchMethod}) associated to the corresponding beam particle. We then apply the ``thin slice" method to measure the ``raw" hadronic cross section (Section \ref{ch:ThinSliceMethod}). A series of corrections are then evaluated and applied to obtain the final cross section (Section \ref{ch:MCCorrections}).

@@

If the particle enters a slice, it contributes to the $N_{\text{Inc}}( E^{kin})$ distribution in the energy bin corresponding to its kinetic energy in that slice. While into the slice, a hadron may or may not interact. If it interacts in the slice, it  contributes also to the $N_{\text{Int}}(E^{kin})$ distribution in the appropriate energy bin; this occurrence corresponds to the end of the hadron tracking. If the hadron does not interact, it will enter the next slice and the interaction evaluation starts again.
The process is applied to all the hadrons in the sample; the cross section as a function of kinetic energy, $\sigma_{TOT}( E^{kin})$ is then evaluated to be proportional to the ratio $\frac{N_{\text{Int}}( E^{kin})}{N_{\text{Inc}}( E^{kin})}$ -- bin by bin ratio.

>>>>

This chapter describes the general procedure employed to measure  total hadronic interaction cross sections on argon in LArIAT.
Albeit with small differences, both the  ($\pi^{-}$,Ar) and (K$^{+}$,Ar) total hadronic cross section measurements rely on the same procedure. We start by selecting the particle of interest using a combination of beamline detectors and TPC information (Section \ref{ch:ParticleSelectionMethod}). We then perform a handshake between the beamline information and the TPC tracking to assure the selection of the correct TPC track (Section \ref{ch:WC2TPCMatchMethod}) associated to the corresponding beam particle. We then apply the ``thin slice" method to measure the ``raw" hadronic cross section (Section \ref{ch:ThinSliceMethod}). A series of corrections are then evaluated and applied to obtain the final cross section (Section \ref{ch:MCCorrections}).

@@

If the particle enters a slice, it contributes to the $N_{\text{Inc}}( E^{kin})$ distribution in the energy bin corresponding to its kinetic energy in that slice.  Within the slice, the hadron may or may not interact. If it interacts in the slice, it  contributes also to the $N_{\text{Int}}(E^{kin})$ distribution in the appropriate energy bin; this occurrence corresponds to the end of the hadron tracking. If the hadron does not interact, it will enter the next slice and the interaction evaluation starts again.
The process is applied to all the hadrons in the sample; the cross section as a function of kinetic energy, $\sigma_{TOT}( E^{kin})$ is then evaluated to be proportional to the ratio $\frac{N_{\text{Int}}( E^{kin})}{N_{\text{Inc}}( E^{kin})}$ -- bin by bin ratio.
